---
title: "Practical 1: basic workflow for two sample MR"
author: "Ting Ye"
subtitle: "SSGG Short Course: A Introduction to Mendelian Randomization, February 12, 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This exercise demonstrates the basic workflow of two-sample summary data Mendelian randomization of LDL-C on CAD.

### 0. Preparation -- install and load package


Download and Install the `TwoSampleMR`, `mr.divw`, `mr.raps` packages: uncomment and run the code below. Load the installed packages.

```{r,message=F}
# library(remotes)
# install_github("MRCIEU/TwoSampleMR") # need to install before load the package
library(TwoSampleMR)

# install_github("tye27/mr.divw") # need to install before load the package
library(mr.divw)

# install_github("qingyuanzhao/mr.raps") # need to install before load the package
library(mr.raps)

```


Run the code below to obtain a data frame of the available GWAS summary datasets in the database.

```{r}
#ao <- available_outcomes()
```

### 1. Find the traits for the GWAS datasets "ieu-a-2" and "ieu-a-7". 


```{r}
#subset(ao, id %in% c("ieu-a-2", "ieu-a-7"))$trait
```


### 2. Select IVs for the exposure from "ieu-a-2" 

Using significance threshold p1 = 5e-08, with default setting in LD clumping. Try changing the p-value threshold to 1e-3, see how many IVs are left.

```{r}
exposure_dat <- extract_instruments("ieu-a-2", p1 = 5e-08, clump = TRUE, r2 = 0.001, kb = 10000)
```

### 3. Extract IVs for the outcome 

```{r}
outcome_dat <- extract_outcome_data(exposure_dat$SNP, "ieu-a-7")
```


### 4. Harmonize the effect sizes

```{r}
dat <- harmonise_data(exposure_dat, outcome_dat)
```


### 5. Run default MR analyses using the TwoSampleMR

```{r}
out <- mr(dat,method_list = c("mr_ivw"))
out
mr_scatter_plot(out, dat)

# Q-statistic 
mr_heterogeneity(dat)
```


### 6. Run dIVW using mr.divw

```{r}
divw.res<-with(dat,mr.divw(beta.exposure, beta.outcome, se.exposure, se.outcome))

#dIVW estimate and SE
round(c(divw.res$beta.hat,divw.res$beta.se),2)

# F-statistic
round(divw.res$condition/sqrt(divw.res$n.IV),2)

```


### 7. Run MR-raps using mr.raps

```{r}
mr.raps_res<-mr.raps(dat,over.dispersion = F)

# point estimate and SE
round(c(mr.raps_res$beta.hat,mr.raps_res$beta.se),2)

```

### 7. Interpretation of the MR estimate

LDL-C is continuous and was standardized prior to GWAS. 
CAD is binary and logistic regression was used in GWAS. Thus, the MR estimate indicates a 57\% (exp(0.45)-1) increase in the odds of CAD per 1-SD increase in LDL-C level.

